Approximation as Differentiation: Notes

### 2025-05-24 main/introduction.tex
Correction.
Minor corrections.

\subsection{Galois Program Slicing}

@@

      % links
      \draw (top) -- (ioi);
      \draw (top) -- (oii);
      \draw (top) -- (iio);
      \draw (ioi) -- (ooi);
      \draw (ioi) -- (ioo);
      \draw (oii) -- (ooi);
      \draw (oii) -- (oio);
      \draw (iio) -- (ioo);
      \draw (iio) -- (oio);
      \draw (ooi) -- (bot);
      \draw (oio) -- (bot);
      \draw (ioo) -- (bot);
    \end{tikzpicture}
  \end{center}
  The output approximation lattice looks like this, where $1$ is the actual data point that was returned in both runs of the program, and $\bot$ indicates that we are approximating this piece of data away:
  \begin{center}
    \begin{tikzpicture}
      \node (top) at (0,0) {$1$};
      \node [below of=top] (bot) {$\bot$};
      \draw (top) -- (bot);
    \end{tikzpicture}
  \end{center}
  These are not the only choices of approximation lattices that we could have taken. For the input, we have chosen a lattice that allows us to ``forget'' (approximate away) numbers in the input, but not any of the other data or the overall structure. However, other choices are also useful. Indeed, one of the aims of this work is to clarify how to choose approximation structure approriate for different tasks by use of type information. We elaborate on this further in Section \todo{XYZ}.

  Galois program slicing associates with each run of the program a Galois connection telling us how the inputs and outputs are related in that run. The backwards portion $\partial (\mathit{query}\,l)_r$ tells us, given an approximation of the output what the least approximation of the input is needed to generate that output. In the case of the two runs considered in this example, if we say we are not interested in the output by feeding in the least approximation $\bot$, then we find that we only need the least approximation of the input:
  \begin{displaymath}
    \partial (\mathit{query}\,l)_r(\bot) = [(\mathsf{a},\bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]
  \end{displaymath}
  for both $l = \mathsf{a}$ and $l = \mathsf{b}$. If instead we take the greatest approximation of the output (i.e., the output ``$1$'' itself), then the two query runs' backwards approximations return different results:
  \begin{displaymath}
    \begin{array}{l}
      \partial (\mathit{query}\,\mathsf{a})_r(1) = [(\mathsf{a},0), (\mathsf{b},\bot), (\mathsf{a},1)] \\
      \partial (\mathit{query}\,\mathsf{b})_r(1) = [(\mathsf{a},\bot), (\mathsf{b},1), (\mathsf{a},\bot)]
    \end{array}
  \end{displaymath}
  Pieces of the input that were {\em not} used are replaced by $\bot$. As we expect, the run of the query with label $\mathsf{a}$ depends on the entries in the database labelled with $\mathsf{a}$, and likewise for the run with label $\mathsf{b}$.

@@

  In a simple query like this, it is easy to work out the dependency relationship between the input and output. However, the benefit of Galois Program Slicing is that it is {\em automatic} for all programs, no matter how complex the relationship between input and output is. Moreover, by changing what we mean by ``approximation'' we can compute a range of different information about a program.
\end{example}

>>>>

\subsection{Galois Program Slicing}
\label{sec:introduction:galois-slicing}

@@

      % links
      \draw (top) -- (ioi);
      \draw (top) -- (oii);
      \draw (top) -- (iio);
      \draw (ioi) -- (ooi);
      \draw (ioi) -- (ioo);
      \draw (oii) -- (ooi);
      \draw (oii) -- (oio);
      \draw (iio) -- (ioo);
      \draw (iio) -- (oio);
      \draw (ooi) -- (bot);
      \draw (oio) -- (bot);
      \draw (ioo) -- (bot);
    \end{tikzpicture}
  \end{center}
  The output approximation lattice looks like this, where $1$ is the actual data point that was returned in both runs of the program, and $\bot$ indicates that we are approximating this piece of data away:
  \begin{center}
    \begin{tikzpicture}
      \node (top) at (0,0) {$1$};
      \node [below of=top] (bot) {$\bot$};
      \draw (top) -- (bot);
    \end{tikzpicture}
  \end{center}
  These are not the only choices of approximation lattices that we could have taken. For the input, we have chosen a lattice that allows us to ``forget'' (approximate away) numbers in the input, but not any of the other data or the overall structure. However, other choices are also useful. Indeed, one of the aims of this work is to clarify how to choose an approximation structure appropriate for different tasks by use of type information. We elaborate on this further in Section \todo{XYZ}.

  Galois program slicing associates with each run of the program a Galois connection telling us how the inputs and outputs are related in that run. The backwards portion $\partial (\mathit{query}\,l)_r$ tells us, given an approximation of the output, what the least approximation of the input is needed to generate that output. In the case of the two runs considered in this example, if we say we are not interested in the output by feeding in the least approximation $\bot$, then we find that we only need the least approximation of the input:
  \begin{displaymath}
    \partial (\mathit{query}\,l)_r(\bot) = [(\mathsf{a},\bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]
  \end{displaymath}
  for both $l = \mathsf{a}$ and $l = \mathsf{b}$. If instead we take the greatest approximation of the output (i.e., the output ``$1$'' itself), then the two query runs' backwards approximations return different results:
  \begin{displaymath}
    \begin{array}{l}
      \partial (\mathit{query}\,\mathsf{a})_r(1) = [(\mathsf{a},0), (\mathsf{b},\bot), (\mathsf{a},1)] \\
      \partial (\mathit{query}\,\mathsf{b})_r(1) = [(\mathsf{a},\bot), (\mathsf{b},1), (\mathsf{a},\bot)]
    \end{array}
  \end{displaymath}
  Pieces of the input that were {\em not} used are replaced by $\bot$. As we expect, the run of the query with label $\mathsf{a}$ depends on the entries in the database labelled with $\mathsf{a}$, and likewise for the run with label $\mathsf{b}$.

@@

  In a simple query like this, it is easy to work out the dependency relationship between the input and output. However, the benefit of Galois Program Slicing is that it is {\em automatic} for all programs, no matter how complex the relationship between input and output. Moreover, by changing what we mean by ``approximation'' we can compute a range of different information about a program.
\end{example}
